---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tcgaCleaneR

<!-- badges: start -->
![R-CMD-check](https://github.com/AbhishekSinha28/tcgaCleaneR/workflows/R-CMD-check/badge.svg)
<!-- badges: end -->

The goal of `tcgaCleaneR` is to provide a user-friendly R package to help Bioinformaticians with easy access to a tool that can perform Data Wrangling and Data Analysis on TCGA Pan Cancer Dataset. The package contains a subset of the original TCGA Breast Cancer Data collected from [TCGA](https://www.cbioportal.org/datasets). The package also contains a detailed set of functionalities that allows user to identify and handle unwanted variations in the TCGA datasets.

## Installation

You can install the development version of tcgaCleaneR from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("AbhishekSinha28/tcgaCleaneR", ref="master")
```

## TCGA Functionality

This is a quick walk trough of the `tcgaCleaneR` functionalities. For the detailed information on the Data Wrangling and Data Analysis functions and arguments in `tcgaCleaneR` package, you can consider looking at the vignette.

At present, TCGA Pan Cancer Datasets supports Cancer Biology for only _**four**_ Cancer types. These four cancer type (TCGA datasets) are *Breast Cancer (BRCA)*, *Lung Cancer (LUAD)*, *Colon Cancer (COAD)* and *Rectum Cancer (READ)*. This implies that *RUV-III analysis* can only be performed for these four cancer types. This is because the RUV-III approach here requires at least one roughly known biologically homogeneous subclass of samples shared across sources of unwanted variation. Similarly, the vector correlation between Biology and PCs can only be viewed for these four Cancer types.

```{r}
library(tcgaCleaneR)
```

# Data

```{r message=FALSE,warning=FALSE}
data("brca.data")
```

```{r message=FALSE,warning=FALSE}
brca.data
```

# Data Wrangling

## Gene Filter

```{r}
filtered.data <- filterGenesByBiotypes(data=brca.data,gene.type=c("protein.coding"))
```

## Removing lowly expressed genes

```{r}
filtered.data1 <- filterLowExprGenes(data=filtered.data,gene_count = 20,sample_size = 200)
```

## Purity Filter - Filter Samples based on Tumor Purity

```{r}
filtered.data2 <- filterSamplesByPurity(data= filtered.data1,purity_cutoff= 0.50)
```

## Library Size Filter

### Determine Library Size

```{r fig.align='center', fig.width=7, fig.height=5}
plotLibSize(data = filtered.data2, plot_type = "Scatterplot")
```

### Filter samples based on library size

```{r}
filtered.data3 <- filterSamplesByLibSize(data = filtered.data2, ls_cutoff = 17.5)
```


# Data Analysis

## Study Design Plot

The idea behind Study Design plot is to present the summarized information about the filtered data set using HeatMaps. 

```{r fig.align='center', fig.width=7, fig.height=5}
plotStudyOutline(data = filtered.data3)
```

## PCA

### Generate PCA

The principal components (in this context also called singular vectors) of the sample × transcript array of log-counts are the linear combinations of the transcript measurements having the largest, second largest, third largest, etc. variation, standardized to be of unit length and orthogonal to the preceding components. Each will give a single value for each sample.

```{r eval=FALSE}
# Is data input for PCA logical
is.logical(filtered.data3)
```

```{r}
pca_data <- computePCA(data = filtered.data3, nPcs = 7, is.log = FALSE)
```

```{r eval=FALSE, echo=FALSE}
# Generated 7 PCs for first two genes
head(pca_data[['HTseq_counts']]$sing.val$u,2)
```

### Plot PCA

Once we have the PCs generated using the PCA function the next step is to visualize those PCs with respect to the sample features like Time, Tissue, Plate etc., to identify any unwanted variation by identifying patterns in the plots by feature.

```{r fig.align='center', fig.width=7, fig.height=5, message=FALSE, warning=FALSE}
library(ggplot2)
library(cowplot)
pca.plot.data <- plotPC(pca.data = pca_data, data = filtered.data3, group = "Time", plot_type = "DensityPlot", pcs.no = c(1,2,3))
```

### PCs correlation with unwanted variations

```{r fig.align='center',message=FALSE, fig.width=7, fig.height=5, warning=FALSE}
library(tidyverse)
corr_data <- plotPCsVar(pca.data = pca_data, data = filtered.data3, type = "purity", nPCs = 7)
corr_data
```

# RUV - III

## Pseudo Replicate Pseudo Sample(PRPS) Map

```{r fig.align='center', fig.width=7, fig.height=5, message=FALSE, warning=FALSE}
#library(tidyverse)
plotPRPS(data = filtered.data3)
```

## PRPS Generation

```{r message=FALSE, warning=FALSE}
sample.info <- as.data.frame(SummarizedExperiment::colData(filtered.data3))
expr.data <- as.matrix(SummarizedExperiment::assay(filtered.data3, 'HTseq_counts')) # gene expression data
sample.info$ls <- colSums(expr.data) # adding library size variable

df9 <- createPRPS(expr.data, sample.info, librarySize = 'ls', batch=c('Year', 'Plates'), biology = 'Subtypes', purity='Purity_singscore',include.ls=T, include.purity=T, minSamplesPerBatchPS = 3, minSamplesForPuirtyPS = 3, 
                 minSamplesForPurityPerBiology = 12, minSamplesForLibrarySizePerBatch = 6,
                 minSamplesForLibrarySizePS = 3)
```

## RUV-III

```{r}
### data input
library(SummarizedExperiment)
### PRPS values
prps.batch <- df9$ps.batch
colnames(prps.batch) <- unlist(lapply(
  colnames(prps.batch),
  function(x) strsplit(x, '-')[[1]][1]
))
prps.ls <- df9$ps.ls
prps.purity <- df9$ps.purity

raw.data <- as.matrix(SummarizedExperiment::assay(filtered.data3, 'HTseq_counts'))
ruv.data <- cbind(raw.data ,prps.batch ,prps.ls, prps.purity )
ruv.data <- t(log2(ruv.data + 1)) # Taking Log 

### replicate matrix
ruv.rep <- ruv::replicate.matrix(row.names(ruv.data))

gene.annot <-  as.data.frame(SummarizedExperiment::rowData(filtered.data3))

### NCG sets - Select House Keeping Genes
ncg.set <- colnames(ruv.data) %in% gene.annot$Gene_symbol[gene.annot$RNAseq_HK == 'yes']
```

```{r}
#library(BiocParallel)
#library(BiocSingular)
df10 <- runRUV_III_PRPS(ruv.data = ruv.data, ruv.rep = ruv.rep, ncg.set = ncg.set, k=1, return.info = TRUE)
```

# Combined Analysis

## Combined data

```{r}
#library(SummarizedExperiment)
gene.annot <-  as.data.frame(SummarizedExperiment::rowData(filtered.data3))

sample.info <-  as.data.frame(SummarizedExperiment::colData(filtered.data3))

ruv.iii.adj <- t(df10$new.ruv.data[1:ncol(raw.data) , ]) # transpose

raw.count <- SummarizedExperiment::assay(filtered.data3, 'HTseq_counts')
raw.count <- log2(raw.count + 1) # Taking Log 
fpkm <- SummarizedExperiment::assay(filtered.data3, 'HTseq_FPKM')
fpkm <- log2(fpkm + 1) # Taking Log 
fpkm.uq <- SummarizedExperiment::assay(filtered.data3, 'HTseq_FPKM.UQ')
fpkm.uq <- log2(fpkm.uq + 1) # Taking Log 
RUV_III <- ruv.iii.adj

```

```{r}
combined_data <- SummarizedExperiment(assays = list(HTseq_counts = raw.count, HTseq_FPKM = fpkm,
                                                HTseq_FPKM.UQ = fpkm.uq, RUV_III = RUV_III),
                                  colData = sample.info,
                                  rowData = gene.annot)

combined_data
```

## PCA on Combined Data
```{r}
df11 <- computePCA(data = combined_data, nPcs = 7, is.log = TRUE)
```

## PCs correlation with unwanted variations in Combined Data

```{r fig.align='center',message=FALSE, fig.width=7, fig.height=5, warning=FALSE}
df12 <- plotPCsVar(pca.data = df11, data = combined_data, type = "purity", nPCs = 7)
df12
```
